---
title: "Script_correlations"
output: html_document
date: "2025-02-24"
---

# CORRELATIONS

Analysis of the correlations between bacterial genera abundances and alpha diversity indexes, and SCFA levels (normalized by bacterial count).

Load of necessary libraries:

```{r}
library(readr)
library(corrplot)
library(reshape2) #order tables
library(ggplot2) #graphs
library(data.table)
library(tidyverse)
library(flextable)
library(kableExtra)
library(knitr)
library(DT) 
library(dplyr)

setwd("")
```

Load csv:

```{r}
#Bacterial genera with SCFAs levels
genera_SCFA <- read.csv("/playground/paper_natasa/Master-Project-Natasa-Mortvanski/01_tidy_data/genera_SCFAs.csv")

#Alpha diversity indexes with SCFAs levels
diversity_SCFA <- read.csv("/playground/paper_natasa/Master-Project-Natasa-Mortvanski/01_tidy_data/01_tidy_data/diversity_SCFAs.csv")
```

## Bacterial genera vs SCFAs

Interactive table with the correlations:

```{r}
#Correlation
genera_cor <- as.data.frame(cor(genera_SCFA[, c(1:3)], genera_SCFA[,c(4:81)], method = "spearman")) %>% t() #traverse the data.frame to make it more legible

#Visualise the table, added an interactive datatable object for the final Rmarkdown. Round to 3 decimals.
round(genera_cor,3) %>% as.data.frame() %>% tibble::rownames_to_column() %>% datatable(rownames = FALSE, filter ="top", class='row-border hover stripe compact', colnames=c('Taxa', 'Acetic Acid','Propionic Acid','Butyric acid'))
```

First we apply the Spearman's correlation, and then we filtrate by:

-   **alpha = 0.01**

-   **Benjamini-Hochberg test**

We apply the Benjamini-Hochberg test. The Benjamini-Hochberg procedure, also known as the False Discovery Rate (FDR) procedure, is a statistical method used in multiple hypothesis testing to control the expected proportion of false discoveries. In many scientific studies or experiments, researchers test multiple hypotheses simultaneously, but when multiple tests are performed, the probability of obtaining at least one false positive result increases leading to an increased overall type I error rate.

The Benjamini-Hochberg procedure addresses this issue by controlling the FDR, which is defined as the expected proportion of false positives among the rejected hypotheses. The calculation of adjusted p-values in the Benjamini-Hochberg procedure involves comparing each individual p-value to a critical value or threshold. The critical value is determined based on the desired false discovery rate (FDR) control.

BH method relies on p-values being uniformly distributed under the null hypothesis. For example, if the p-values are uniformly distributed, then the p-value 5% of the way down the sorted list should be \~0.05. Accordingly, the procedure consists of sorting the p-values in ascending order, and then dividing each observed p-value by its percentile rank to get an estimated FDR. In this way, small p-values that appear far down the sorted list will result in small FDR estimates, and vice versa.

```{r}
#BENJAMINI-HOCHBERG
#Spearman's method
matrix_genera <- cor(genera_SCFA, method = "spearman") %>% t()
test_genera <- cor.mtest(genera_SCFA, conf.level = 0.95) 

#cor.mtest returns three objects (p, lowCI, uppCI). We keep p-values.
pvalue_genera <- as.data.frame(test_genera$p)

#We create a dataframe with the name of the taxon as a row.name 
df <- tibble::rownames_to_column(pvalue_genera, "taxon") %>% melt()
names(df) <-c('taxon','variable','p')

#Ajust the p value acording to the Benjamini-Hochberg method
df <- cbind(df, p.adj.bh=p.adjust(df$p, method='BH'))
df$p <-NULL

p_value_bh <- reshape2::dcast(df, taxon ~ variable)
#We keep the sample info in the row.name
rownames(p_value_bh) <- p_value_bh$taxon 
p_value_bh$taxon <- NULL

#We ensure the matrix of the correlations and the p values are the same order 
ind = colnames(matrix_genera)
corr = matrix_genera[ind, ind]
p = p_value_bh[ind, ind]

#Now we filtrate the results: first the columns of interest and then according to p values 
corr_genera <- as.matrix(corr[, c(1:3)], corr[, c(4:81)])

#We delate first rows (they are the SCFAs ans we don't need them), to have the same length
corr_genera <- corr_genera[-(1:3),]
p_genera_BH <- p %>% slice(1:3) %>% select(4:81) %>% t()

#We create a dataframe with the correlation results and the p-values
genera_SCFA_BH <- data.frame(corr_genera, p_genera_BH)
colnames(genera_SCFA_BH)[4] <- 'acetic_pvalue_BH'
colnames(genera_SCFA_BH)[5] <- 'propionic_pvalue_BH'
colnames(genera_SCFA_BH)[6] <- 'butyric_pvalue_BH'

colnames(genera_SCFA_BH)[1] <- 'Acetate'
colnames(genera_SCFA_BH)[2] <- 'Propionate'
colnames(genera_SCFA_BH)[3] <- 'Butyrate'

#We filtrate results according to the p values 
genera_SCFA_BH_signif <- filter(genera_SCFA_BH, genera_SCFA_BH$acetic_pvalue_BH < 0.01 |
                               genera_SCFA_BH$propionic_pvalue_BH < 0.01 |
                               genera_SCFA_BH$butyric_pvalue_BH < 0.01)

#Now we create a matrix of the results (correlations) and a dataframe of the p values 
matrix_genera_BH <- as.matrix(genera_SCFA_BH_signif[1:3]) %>% t()
p_BH <- as.data.frame(genera_SCFA_BH_signif[4:6]) %>% t()
rownames(p_BH) <- c('Acetate','Propionate','Butyrate')
```

### Figure 5A. Correlation plot (genera)

Plot:

```{r}
#PLOT GENERA VS SCFAs
plot_genera_BH <- corrplot(matrix_genera_BH, method="circle", type="full", col = COL2(c("RdBu")), tl.col = "black", tl.srt = 45, tl.cex = 1.5,cl.cex = 1.5, p.mat = data.matrix(p_BH), insig = "blank", sig.level = 0.01)

dev.copy(jpeg,filename="genera_corrplot.jpg")
dev.off ()
```

## Diversity indexes vs SCFAs

Interactive table with the correlations:

```{r}
#Correlation
diversity_cor <- as.data.frame(cor(diversity_SCFA[, c(1:3)], diversity_SCFA[, c(4:13)], method = "spearman")) %>% t()

#Visualise the table, added an interactive datatable object for the final Rmarkdown. Round to 3 decimals.
round(diversity_cor,3) %>% as.data.frame() %>% tibble::rownames_to_column() %>% datatable(rownames = FALSE, filter ="top", class='row-border hover stripe compact', colnames=c('Taxa', 'Acetic Acid','Propionic Acid','Butyric acid'))
```

First we apply the Spearman's correlation, and then we filtrate by p-value and BH:

```{r}
#BENJAMINI-HOCHBERG
#Spearman's method
#BENJAMINI-HOCHBERG
#Spearman's method
matrix_diversity <- cor(diversity_SCFA, method = "spearman") %>% t()
test_diversity <- cor.mtest(diversity_SCFA, conf.level = 0.95) 

#cor.mtest returns three objects (p, lowCI, uppCI). We keep p-values.
pvalue_diversity <- as.data.frame(test_diversity$p)

#We create a dataframe with the name of the taxon as a row.name 
df <- tibble::rownames_to_column(pvalue_diversity, "taxon") %>% melt()
names(df) <-c('taxon','variable','p')

#Ajust the p value acording to the Benjamini-Hochberg method
df <- cbind(df, p.adj.bh=p.adjust(df$p, method='BH'))
df$p <-NULL

p_value_bh <- reshape2::dcast(df, taxon ~ variable)
#We keep the sample info in the row.name
rownames(p_value_bh) <- p_value_bh$taxon 
p_value_bh$taxon <- NULL

#We ensure the matrix of the correlations and the p values are the same order 
ind = colnames(matrix_diversity)
corr = matrix_diversity[ind, ind]
p = p_value_bh[ind, ind]

#Now we filtrate the results: first the columns of interest and then according to p values 
corr_diversity <- as.matrix(corr[, c(1:3)], corr[, c(4:81)])

#We delate first rows (they are the SCFAs ans we don't need them), to have the same length
corr_diversity <- corr_diversity[-(1:3),]
p_diversity_BH <- p %>% slice(1:3) %>% select(4:13) %>% t()

#We create a dataframe with the correlation results and the p-values
diversity_SCFA_BH <- data.frame(corr_diversity, p_diversity_BH)
colnames(diversity_SCFA_BH)[4] <- 'acetic_pvalue_BH'
colnames(diversity_SCFA_BH)[5] <- 'propionic_pvalue_BH'
colnames(diversity_SCFA_BH)[6] <- 'butyric_pvalue_BH'

colnames(diversity_SCFA_BH)[1] <- 'Acetate'
colnames(diversity_SCFA_BH)[2] <- 'Propionate'
colnames(diversity_SCFA_BH)[3] <- 'Butyrate'

#We filtrate results according to the p values 
diversity_SCFA_BH_signif <- filter(diversity_SCFA_BH, diversity_SCFA_BH$acetic_pvalue_BH < 0.01 |
                               diversity_SCFA_BH$propionic_pvalue_BH < 0.01 |
                               diversity_SCFA_BH$butyric_pvalue_BH < 0.01)

#Now we create a matrix of the results (correlations) and a dataframe of the p values 
matrix_diversity_BH <- as.matrix(diversity_SCFA_BH_signif[1:3]) %>% t()
p_BH <- as.data.frame(diversity_SCFA_BH_signif[4:6]) %>% t()
rownames(p_BH) <- c('Acetate','Propionate','Butyrate')
```

### Figure 5B. Correlation plot (diversity)

Plot:

```{r}
#Ajust names
colnames(matrix_diversity_BH) <- c("Shannon entropy", "Chao1", "Fisher alpha", "Margalef", "Gini index", "Menhinick", "Simpson", "Faith PD", "Pielou evenness")
colnames(p_BH) <- c("Shannon entropy", "Chao1", "Fisher alpha", "Margalef", "Gini index", "Menhinick", "Simpson", "Faith PD", "Pielou evenness")
rownames(p_BH) <- c('Acetate', 'Propionate', 'Butyrate')

#PLOT ALPHA-DIVERSITY VS SCFAs
plot_diversity_BH <- corrplot(matrix_diversity_BH, method="circle", type="full", col = COL2(c("RdBu")), tl.col = "black", tl.srt = 45, tl.cex = 1, p.mat = data.matrix(p_BH), insig = "blank", sig.level = 0.01)

dev.copy(jpeg,filename="diversity_corrplot.jpg")
dev.off ()
```
